 

### 2021-05-11 vertical-thermal-review.Rmd
7500 words total without remote sensing, modeling and conclusion

*[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]*

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), as optimum temperature $T_{opt}$ for $J_{max}$ has shown to increase with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020]. Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun and shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical profile in tropical forests [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].

@@

*(KAT: let's cut the portion on T50 down to one or at most two par:)*

@@

Across forest species, typical values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al. 2017). $T_{50}$ and $T_{crit}$ decreases with elevation, as seen in tropical forests in Panama and Columbia and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017] and varies with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020]. Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$. On the other hand, leaves of species with greater construction costs, observed through LMA, had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020]. However, higher photosynthetic heat tolerance does not necessary increase the breadth of leaf thermal safety margins. Rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had narrower breadth of thermal safety margins (difference between maximum leaf temperature and $T_{50}$) [@perezPhotosyntheticHeatTolerances2020]

@@

In other cases, leaf thermal tolerance reflects microclimatic stress along a vertical gradient. For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions. Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [@curtis_intracanopy_2019].

@@

In two forests, north-eastern oak dominated-deciduous forest in the US and conifer-dominated temperate New Zealand rainforest, across species $Q_{10}$ was found to be greater in upper-canopy leaves than in lower-canopy leaves, where values differed between species [@turnbullScalingFoliarRespiration2003]. Similarly, in sweetgum (*Liquidambara styraciflua*), upper canopy sun leaves showed greater $Q_{10}$ than lower within-canopy shade leaves [@harleyEffectsLightTemperature1996]. However, in other feild studies, $Q_{10}$ didnot vary across species vertical profiles and light exposure such as in a lowland tropical Australian forest study [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta [@bolstad_foliar_1999] $Q_{10}$. At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*). @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.

With temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum. Beyond this inflection point, fluxnet global data analysis shows that photosynthesis declines whereas respiration increases exponentially, therefore, moving from carbon sink to carbon source with increasing temperature, independent of water and irradiance [@duffyHowCloseAre2021]

@@

**Isoprene production**
- high emission rates occur in the mid-canopy, and even from very small statured trees and shrubs, at a site in the Brazilian Amazon (Taylor et al., in prep).
- The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
-  Within species, isporene production scales with light/ T; however regardless of light, isoprene emission increases with T such as in understory shaded leaves?

@@

More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
Many observations of larger trees suffering more during drought (Bennett et al. 2015).
One likely mechanism behind this pattern is that the crowns of larger trees exist higher in the vertical profile and therefore are exposed to higher evaporative demand and solar radiation [@mcgregor_tree_nodate].
Although drought is primarily a hydraulic problem, lack of water --> lower gs --> higher leaf T, so leaves face tradeoffs of water loss vs potentially damaging leafT.

On the basis of phenology, temperate understory  trees tend to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al. 2013], and, increase in annual temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renner, 2019]. However, short-term May spring heat waves have shown to negetively impact partially-expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis in reflushed leaves, more than in understory saplings that are buffered from such conditions [Filewood and Thomas, 2013].

@@

Additionally, increased exposure to light increases non structural carbon reserve (NSC), plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availability become limited, however with available water overall sensitivity to stress decreases. Understory shaded/younger trees have comparatively less NSC reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotranpiration demands, as well as greater investment in light harvesting than in photosynthesis--with greater overall sensitivity to stress and reduced tolerance. Larger NSC reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature]. Thus, overall stress sensitivity decreases and tolerance increases with increasing height, light and age.[@niinemetsResponsesForestTrees2010a]

@@

Interaction between degrees of shade x ontogeny with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.

@@

*Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020]. Larger trees suffer disproportionally greater hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020]. However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [@perezIncreasingHumidityThreatens2018].*

*While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020]. Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020]. Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020]. Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]. Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020]. In addition, understory growth phenology is more sensitive to warming than canopy-- showing earlier spring onset and later autumn senescense [@zohnerOngoingSeasonallyUneven2019].*

*The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a]. Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019]. In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humidity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [@tibbittsHumidityPlants1979; @perezIncreasingHumidityThreatens2018].*

 if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered

@@

Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a]. Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020]. Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020]. Each of these changes have potential feedbacks to climate change.

increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness

>>>>

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that optimum temperature $T_{opt}$ of $J_{max}$ increased with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020].Similar results were shown for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun and shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].

@@

@@

Typical $T_{50}$ values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al. 2017). With elevation, $T_{50}$ and $T_{crit}$ decrease, as seen in tropical foress and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], and vary with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020]. Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$. On the other hand, leaves of species with greater construction costs, greater LMA, had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020]. However, higher photosynthetic heat tolerance does not necessary increase the breadth of leaf thermal safety margin. Rather, across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins (difference between maximum leaf temperature and $T_{50}$) [@perezPhotosyntheticHeatTolerances2020]

@@

In other cases, leaf thermal tolerance reflects microclimatic stress along a vertical gradient. For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions. Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, low-wind speed, $T_{air} and greater canopy [@curtis_intracanopy_2019].

@@

Where $Q_{10}$ is greater in exposed upper-canopy leaves than in shaded lower-canopy leaves in across species in north-eastern oak deciduous forest, conifer temperate rainforest and in *Liquidambara styraciflua* tree, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996] but in other feild studies $Q_{10}$ doesnot vary across species vertical profiles in a lowland tropical Australian forest study [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta, with a slight tendency to increase down the canopy relative to upper-canopy leaves as T increases (*see their Fig. 1*)[@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.

Additionally, a fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]

@@

**Isoprene production**
High emission rates occur in the mid-canopy, and even from very small statured trees and shrubs, at a site in the Brazilian Amazon (Taylor et al., in prep).
The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564). Within species, isporene production scales with light/ T; however regardless of light, isoprene emission increases with T such as in understory shaded leaves?

@@

More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality. Biophysically, greater vapor pressure deficit, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al. 2016]. On the other hand, greater relative humidity can do the opposite, where it stifles evapotranspiration thereby increasing leaf temperatures, particularly in canopy tops where irradiance can have a greater effect [Fig.3;@perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. However, soil water loss in conditions of drought effect larger trees more than small trees in less than a few days. Primarily because, greater biophysical evaporative demand in the canopy cannot meet long-distance water transport from the ground due to limited soil moisture [Norway spruce study, bennett; @mcgregor_tree_nodate].

On the basis of phenology, because understory trees leaf out earlier than canopy due to ontogenetic variation in tree size [Vitasse et al. 2013], understory growth is more sensitive to annual warming than canopy - showing earlier spring onset and later autumn senescense, particularly in temperate deciduous forests [Zohnner and Renner, 2019]. However, late spring short-term heat waves negatively impact partially- expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis in reflushed leaves, more than understory saplings that are buffered from such conditions [Filewood and Thomas, 2013].

If we expect 'shade' plants to have a different heat sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered.

@@

Additionally, interacting biophysical and biological stressors show that stress sensitivity is greater in buffered small trees, and decreases where tolerance increases with height, light with available water.  Partly because large NSC reserves in taller trees due to available light allow them to tolerate greater number of stressors than buffered trees that are light and NSC limited [@niinemetsResponsesForestTrees2010a]

@@

@@

*While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020]. Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020]. Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020]. Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]. Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020]. Additionally, warming and heat waves can alter canopy and understory phenology and forest photosynthesis [@zohnerOngoingSeasonallyUneven2019].*

@@

*The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020]. With larger trees suffering disproportionally greater hydraulic stress during drought, can contribute to increasing larger tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020], or increase in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughnessare, also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].*

*Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Open-forest might experience greater microclimate warming than forests, having implications for forest regeneration, particularly in high-altitude regions [@vonarxSpatiotemporalEffectsForest2012a]. Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020]. Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020]. Each of these changes have potential feedbacks to climate change.*
